The Rockerverse: Images, Packages, and Applications for Containerization with R

### 2020-01-23 manuscript.Rmd
update geospatial use case, move r-ci to testing section

## Development and debugging
\label{development}

@@

First, such containers can greatly facilitate **fixing bugs**, because developers can readily start and later discard a container to investigate a bug report without affecting their regular system, and potentially having to revert changes back.
Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
\citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).

Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis. For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and therefore the maintainers spend a lot of efforts not to break these.
```{r geospatial_deps, eval=FALSE, include=FALSE}
length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
```
Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf. \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.

@@

Fifth, Docker images can be used **on CI platforms** to streamline the testing of packages.
\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
\citet{cardozo_faster_2018}, also on Travis&nbsp;CI saved time by re-using the testing image as the base iamge for an image intended for publication on Docker&nbsp;Hub.
Especially for long running tests or complex system dependencies, these approaches seems to improve the development.
Not due to a concern about time, but to control the environment used on a CI server, even this manuscript was rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository).
This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.

>>>>

## Development, debugging, and testing
\label{development}

@@

First, such containers can greatly facilitate **fixing bugs and quick evaluation**, because developers and users can readily start and later discard a container to investigate a bug report or try out a piece of software \citep[cf.][]{ooms_opencpu_2017}.
The container does not affect their regular system.
Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for cusotmization during software evaluation.
\citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).

Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis. For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
GDAL is a system dependen^cy allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and therefore the maintainers spend a lot of efforts not to break these.
Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf. \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
```{r geospatial_deps, eval=FALSE, include=FALSE}
length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
```

@@

Fifth, Docker images can be used **on CI platforms** to streamline the testing of packages.
\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
\citet{cardozo_faster_2018} saved time, also on Travis&nbsp;CI, by re-using the testing image as the base for an image intended for publication on Docker&nbsp;Hub.
[`r-ci`](https://github.com/ColinFay/r-ci) is in turn used with [GitLab&nbsp;CI](https://docs.gitlab.com/ee/ci/), which itself is built on top of Docker images: the user specifies a base Docker image, and the whole tests are run inside this environment.
The `r-ci` image stack combines `rocker` versioning and a series of tools specifically designed for testing in a fixed environment with a customized list of preinstalled packages.
Especially for long running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
Not due to a concern about time, but to control the environment used on a CI server, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website with every change (see `.travis.yml` and `Dockerfile` in the manuscript repository).
This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.
